Grove: A Convergent Collaborative Structure-Editor Calculus

### 2022-08-09 introduction.tex
working on intro

TODO: fix the following warning which are reported by LaTeX but not Overleaf: Class acmart Warning: A possible image without description on input line 321.

>>>>

A \emph{collaborative structure editor} is an editor that
(1) allows multiple concurrent users to work on a shared document, while also
(2) providing structure-aware editor services such as projectional editing, syntax highlighting, or automatic code folding.
%
Collaborative editing research focuses on the design and implementation of real-time, multi-user, character-based communication systems,
whereas structure editors typically presume a more complex document schema and then focus on some other aspect of the user experience.
In both settings, preservation of user intent is a core technical challenge.
%
Although collaborative editors and structure editors have overlapping goals (optimal user experience)
and complementary design challenges (subject-subject versus subject-object harmony),
to our knowledge, there is no comprehensive, principled account of their combined use.

Since collaborative editors are essentially distributed systems, existing work tends to focus on extensions to distribution protocols.
Lots of examples using OT. (Google Docs)
OT is complex and largely textual.
OT can make sense for real time systems: users typically change one character at a time, and instant feedback can help to prevent conflicts.
On the other hand, OT system designs can be difficult to extend.

Alternatively, there's CRDT. (Peritext?)
CRDT is easier to implement, but harder to design for.
There's an All-CRDT editor---it turned out to be not so realistic. (what's it called again?)

Structure editing has been a recurring theme in the computer science literature since at least Engelbart's ``Mother of All Demos.''
Provides automation for domain experts and reduces the barrier to entry for everyone else.
Popular for editing programs, i.e., for programming language-specific editors.

However, modern program editors typically disable editor services (like what?) when the document is not in a consistent state,
a phenomenon called the ``gap problem.''
Of course, in the presence of multiple concurrent users, the problem gets worse.

In a collaborative setting,
Hazel is a structure editor with support for advanced editing services, (e.g., semantic actions?).

\newpage
